\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-24 chapters/wcble.tex
Set manuscript draft

\todo{comments liesel}

This chapter, except Sections~\ref{sec:wcble/intro},
\ref{sec:wcble/literature/single-comp} and
\ref{sec:wcble/literature/multi-comp}
\todo{properly assess which paragraphs were published and which not}

@@

\subsection{Contribution}

@@

\section{Methods}

\subsection{Classifier-based Latency Estimation}
\label{sec:cble}

@@

\ac{cble}, summarized in Algorithm~\ref{alg:cble}, works by training a
\textit{first-stage} classifier $\mathcal{C}(\theta,f)$
defined within a time period $[s_1:s_2]$ with a set of parameters $\theta$ and a
decision function $f(\mat{X}[:,s_1:s_2],\theta) \rightarrow
y$ outputting a classification score $y\in\mathbb{R}$ for a given epoch $\mat{X}$,
such that
\begin{equation}
  \theta = \mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
  \label{eq:cble-train}
\end{equation}
Then,
%instead of applying the classifier directly to an unseen, cropped
%testing epoch $\mat{X}[:,s_1:s_2]$ and obtaining a single score value $y$,
$f$ can be applied to all (possibly overlapping) slices of length $s_2-s_1$ of
an epoch $\mat{X}$, resulting in a vector of score values
$\mathbf{y}=[y_1\,\ldots\,y_R]^T \in\mathbb{R}^R$ such that
\begin{equation}
  y_s = f(\mat{X}[:,s:s+(s_2-s_1)],\theta)\quad \forall s\in 1,\ldots,R
	\label{eq:score}
\end{equation}
with $R = S-(s_2-s_1)$.
To leverage \ac{cble} for \ac{erp} classification, the score vectors $\mathbf{y}$ can be
arranged in matrices $\mat{Y}^\mathrm{train}\in\mathbb{R}^{N\times R}$ and $\mat{Y}^\mathrm{test}\in\mathbb{R}^{M\times R}$.
These can be further classified by training a \textit{second-stage} classifier on
$\mat{Y}^\mathrm{train}$ and class labels $\mathbf{l}^\mathrm{train}$.
However, the resulting score-over-time vectors per epoch still suffer from jitter.
For classification, we follow the approach of \cite{Mowla2017}, using a
maximum-level hierarchical Daubechies-4 wavelet transform to reduce
dimensionality before classification with the second-stage classifier.
In the \ac{cble}-decoder, it is this wavelet transform that decreases the sensitivity
to latency differences, actively compensating for \ac{erp} latency jitter.

@@

\begin{algorithm}[H]
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l},
		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
	\begin{algorithmic}[1]
		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
    N$
      %\algorithmiccomment{Train stage 1}
		\Repeat
		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l}$
		\For{$n=1\ldots N$}
    \State  $s_n \gets
			    \mathrm{median}\left[Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
		\EndFor
    \Until{convergence or maximum iterations reached}
    \For{$n=1\ldots N$}%\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets
      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
    \EndFor

@@

We compare our proposed \ac{wcble} algorithm to \ac{cble} on both a latency
estimation task and a binary classification task with simulated data.
Four settings are evaluated to assess both the impact of the iterative Woody
procedure and the extraction of robust latency features.
These are CBLE-median and WCBLE-median, where the median of the latency
probability distribution is used as latency estimate, and CBLE-peak and
WCBLE-peak, where the maximum of the score output is used as latency estimate.

\subsection{Data simulation}
To verify our approach, we simulated 32 channel \ac{eeg} data with standard
10-20 electrode positions using the \texttt{simulate\_evoked} method of the
MNE (version 1.8)~\cite{Gramfort2013} software package.
\Acp{erp} were generated by projecting a source time course in a dipole in the
left hemisphere to the scalp electrodes using a boundary element method forward
model~\cite{Mosher1999} MNE's \texttt{fsaverage} source space and anatomy.
Pink temporal noise is generated by passing gaussian noise through an infinite
impulse response filter with transfer function
$\frac{B(z)}{A(z)} = \frac{1}{1 -1z^{-1}+0.15z^{-2}}$.
Pink spatial noise is also added, using a noise covariance matrix constructed from the
cosine distances between electrode positions and scaling it by the inverse of
the resulting spatial.
We refer to~\textcite{Gramfort2014} for implementation details.

Evoked potentials are simulated for three different source time course waveforms,
a gaussian, a morlet wavelet and a sinusoid pulse with variable latency,
defined by the following formulas:
\begin{equation}
  f_\text{gaussian}(t,l) = a
  \frac{1}{\varsigma\sqrt{2\pi}}e^{-\frac{1}{2}\frac{\left(t-l\right)^2}{\varsigma/2}}
\end{equation}
\begin{equation}
  f_\text{morlet}(t,l) = ae^{-\left(\frac{t-l}{\frac{c}{2\pi f}}\right)^2}
\end{equation}
\begin{equation}
  f_\text{sine}(t,l) =
  \begin{cases}
    a\sin\left(2\pi f\left(t-l\right)\right) & \text{if  $l-\frac{1}{2f} < t < l-\frac{1}{2f}$} \\
    0 & \text{otherwise}
  \end{cases}
\end{equation}
with time $t$, latency $l$, and amplitude $a=\num{1e{-7}}$V.
The standard deviation that determines the width gaussian $\varsigma=0.1s$.
For the morlet wavelet, the number of cycles $c=3$ and the frequency $f=4Hz$.
For the sine pulse, the frequency $f=4$
For each of these, 100 target epochs containing the waveform and noise and 100
non-target epochs containing only noise were simulated.
Additionally, the target epochs were jittered by setting the latency offset $l$
of each target epoch to a random latency drawn from a normal distribution with 0 mean and standard
deviations of respectively $\sigma=0.1,0.2,0.3$s.
The noise was also scaled at different 32  different noise levels, with the \ac{snr} of a
target epoch ranging from 0 to -31dB, with
\begin{equation}
  \text{SNR} = 10\log_{10}\frac{\text{var}f(t,0)}{\text{var}\left(\text{noise}\right)}
\end{equation}

A sample of the simulated evoked data is displayed in
Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
\ref{fig:wcble/sim-sine}.
\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
\fullpagefig{figures/wcble/simulated-sine.pdf}{}{fig:wcble/sim-sine}
\todo{Properly align plots and label axes}

@@

CV/LOOCV?
\subsection{Classification}
refer to later chapter for use on real erp data

>>>>

\todo{comments liesel}
Sections~\ref{sec:wcble/literature/contrib} were published \ref{sec:wcble/methods}
by~\textcite{VanDenKerchove2024}

@@

\subsection{Contribution}
\label{sec:wcble/literature/contrib}

@@

\section{Methods}
\label{sec:wcble/methods}

\subsection{Classifier-based Latency Estimation}
\label{sec:wcble/methods/cble}

@@

\Ac{cble}, summarized in Algorithm~\ref{alg:cble}, works by training a
\textit{first-stage} classifier $\mathcal{C}(\theta,f)$
defined within a time period $[s_1:s_2]$ with a set of parameters $\theta$ and a
decision function $f(\mat{X}[:,s_1:s_2],\theta) \rightarrow
y$ outputting a classification score $y\in\mathbb{R}$ for a given epoch $\mat{X}$,
such that
\begin{equation}
  \theta = \mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
  \label{eq:cble-train}
\end{equation}
Then,
%instead of applying the classifier directly to an unseen, cropped
%testing epoch $\mat{X}[:,s_1:s_2]$ and obtaining a single score value $y$,
$f$ can be applied to all (possibly overlapping) slices of length $s_2-s_1$ of
an epoch $\mat{X}$, resulting in a vector of score values
$\mathbf{y}=[y_1\,\ldots\,y_R]^T \in\mathbb{R}^R$ such that
\begin{equation}
  y_s = f(\mat{X}[:,s:s+(s_2-s_1)],\theta)\quad \forall s\in 1,\ldots,R
	\label{eq:score}
\end{equation}
with $R = S-(s_2-s_1)$.
To leverage \ac{cble} for \ac{erp} classification, the score vectors $\mathbf{y}$ can be
arranged in matrices $\mat{Y}^\mathrm{train}\in\mathbb{R}^{N\times R}$ and $\mat{Y}^\mathrm{test}\in\mathbb{R}^{M\times R}$.
These can be further classified by training a \textit{second-stage} classifier on
$\mat{Y}^\mathrm{train}$ and class labels $\mathbf{l}^\mathrm{train}$.
However, the resulting score-over-time vectors per epoch still suffer from jitter.
For classification, we follow the approach of \cite{Mowla2017}, using a
maximum-level hierarchical Daubechies-4 wavelet transform to reduce
dimensionality before classification with the second-stage classifier.
In the \ac{cble}-decoder, it is this wavelet transform that decreases the sensitivity
to latency differences, actively compensating for \ac{erp} latency jitter.

@@

\begin{algorithm}[H]
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l})
		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
	\begin{algorithmic}[1]
		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
    N$
      %\algorithmiccomment{Train stage 1}
		\Repeat
		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l}$
		\For{$n=1\ldots N$}
    \State  $s_n \gets
			    \mathrm{median}\left[Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
		\EndFor
    \Until{convergence or maximum iterations reached}
    \For{$n=1\ldots N$}%\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets
      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
    \EndFor

@@

%Four settings are evaluated to assess both the impact of the iterative Woody
%procedure and the extraction of robust latency features.
%These are CBLE-median and WCBLE-median, where the median of the latency
%probability distribution is used as latency estimate, and CBLE-peak and
%WCBLE-peak, where the maximum of the score output is used as latency estimate.

\todo{
refer to later chapter for use on real erp data
}
\subsection{Data simulation}
To verify our approach, we simulated 16 channel \ac{eeg} data with standard
10-20 electrode positions using the \texttt{simulate\_evoked} method of the
MNE (version 1.8)~\cite{Gramfort2013} software package.
\Acp{erp} were generated by projecting a sine wave pulse source time course in a dipole in the
left hemisphere to the scalp electrodes using a boundary element method forward
model~\cite{Mosher1999} MNE's \texttt{fsaverage} source space and anatomy.
Pink temporal noise is generated by passing gaussian noise through an infinite
impulse response filter with transfer function
$\frac{B(z)}{A(z)} = \frac{1}{1 -1z^{-1}+0.15z^{-2}}$.
Pink spatial noise is also added, using a noise covariance matrix constructed from the
cosine distances between electrode positions and scaling it by the inverse of
the resulting spatial.
We refer to~\textcite{Gramfort2014} for implementation details.
%Evoked potentials are simulated for three different source time course waveforms,
%a gaussian, a morlet wavelet and a sinusoid pulse with variable latency,
%defined by the following formulas:
%\begin{equation}
%  f_\text{gaussian}(t,l) = a
%  \frac{1}{\varsigma\sqrt{2\pi}}e^{-\frac{1}{2}\frac{\left(t-l\right)^2}{\varsigma/2}}
%\end{equation}
%\begin{equation}
%  f_\text{morlet}(t,l) = ae^{-\left(\frac{t-l}{\frac{c}{2\pi f}}\right)^2}
%\end{equation}
The source time course is defined by the following function:
\begin{equation}
  s(t,l) =
  \begin{cases}
    a\sin\left(2\pi f\left(t-l\right)\right) & \text{if  $l-\frac{1}{2f} < t < l-\frac{1}{2f}$} \\
    0 & \text{otherwise}
  \end{cases}
\end{equation}
over time $t$, with latency $l$ and amplitude $a=\num{1e{-7}}$V and frequency
$f=4$.
%The standard deviation that determines the width gaussian $\varsigma=0.1s$.
%For the morlet wavelet, the number of cycles $c=3$ and the frequency $f=4Hz$.
%For the sine pulse, the frequency $f=4$

%For each of these,
100 target epochs containing the evoked potential and noise and 100
non-target epochs containing only noise were simulated.
The target epochs were jittered by setting the latency offset $l$
of each target epoch to a random latency drawn from a normal distribution with 0 mean and standard
deviations of respectively $\sigma=0.1,0.2,0.3$s.
The noise was scaled at different 32  different noise levels, with the \ac{snr} of a
target epoch ranging from 0 to -31dB, with
\begin{equation}
  \text{SNR} = 10\log_{10}\frac{\text{var}s(t,0)}{\text{var}\left(\text{noise}\right)}
\end{equation}

%A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-sine}.
%\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
%\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
\fullpagefig{figures/wcble/simulated-sine.pdf}{}{fig:wcble/sim-sine}
\todo{Properly align plots and label axes}

@@

We compared our proposed \ac{wcble} algorithm to \ac{cble} on a latency
estimation task using the pearson correlation coefficient $\rho$\todo{equation} at different
jitter and \ac{snr} levels.
For both \ac{cble} and \ac{wcble}, target latencies where extracted in a
10-fold cross-validation scheme using the robust median latency method after
fitting the model.
\Ac{wcble}'s maximum iteration number was set to 64.
Results are presented in Figure~\ref{fig:wcble/results/latency}.

\begin{figure}
  \includegraphics[width=\textwidth]{figures/wcble/fig_latency.pdf}
  \caption{align results}
  \label{fig:wcble/results/latency}
\end{figure}
\todo{invert x axis}
\todo{caption}
\todo{properly label axes}
\todo{discuss results}

\subsection{Classification}
Similar to the previous experiment, classification accuracy between target and
non-target epochs is assessed.

\todo{insert results figure and discuss}
